An FPGA Implementation of a Long Short-Term Memory Neural Network

### 2016-10-12 paper/paper.tex
8 pages only

Hardware platforms could achieve a considerable speed-up over software implementations, which would prove
useful for high data throughput systems, where the calculation overhead is critical and limits the performance.
Furthermore, a hardware implementation can even benefit \emph{offline} Deep Learning tasks by providing the results
of a given experiment faster than in a regular CPU, yielding an increase in scientific productivity.

@@

Summarizing, this article begins by explaining what are LSTM networks -- and their mathematical
details -- in Section~\ref{sec:lstmnn}, followed by a quick overview of their state
of the art in Section~\ref{sec:relwork}. Section~\ref{sec:proparch} outlines the proposed
hardware architecture and its main constituent modules, and the performance and synthesis
results are reported in Section~\ref{sec:results}. Concluding remarks are given in
Section~\ref{sec:concl}.

@@

\begin{enumerate}
    \item Multiply matrix $\mb{W}_*$ by the input vector $\mb{x}^{(t)}$
    \item Multiply matrix $\mb{R}_*$ by the previous layer output vector $\mb{y}^{(t-1)}$
    \item Sum the bias vector $\mb{b}_*$ to the remaining matrix-vector dot product results.
\end{enumerate}
Assuming that the network size $N$ is always larger than the input size $M$, if we use the matrix-vector dot product units of Section~\ref{sec:proprarch_dot}, the multiplication
in task 1 takes approximately $K_G\cdot M$ cycles and the one in task 2 takes $K_G\cdot N$ cycles. This way, tasks 1 and 2 can be performed in parallel, and we can use the extra
time that task 2 takes, relative to task 1,  to perform task 3, and sum the bias vector to the output of task 1, whose result is ready by that time. The module is triggered by a
\verb+beginCalc+ input signal that activates the internal state machine, and outputs a \verb+dataReady+ signal that informs the network that the calculations have been concluded.
After validating and simulating this block, and taking into account the internal state-machine and that the internal datapath is pipelined, the exact number
of clock cycles this module takes to produce an output is $6+K_G\cdot N$.

@@

Furthermore, we can avoid a naive translation of the Equations~\ref{eq:equationsLSTM}, which would replicate unnecessary resources (such as elementwise multipliers
and activation function calculators) and require more area to save a negligible number of clock cycles, by noting that one of the operands is the output
of a $\tanh(\mb{x})$ block and the other of a $\sigma(\mb{x})$, and they are then multiplied (elementwise, of course) together. Instead of replicating
these `$\tanh$-$\sigma$-$(\cdot)$wise' structures, we use a \emph{single one} and choose the operand according to the state that the network is currently in. The
issue about the elementwise multiplier for $\mb{c}^{(t)}$, which does not use the $\tanh$ activation function, can be solved by adding another multiplexer that chooses
between the output of the $\tanh(\mb{x})$ module or the signal $\mb{c}^{(t-1)}$. These ideas resulted in the LSTM network design of Figure~\ref{fig:network-opt}, which
is mathematically equivalent to Equations~\ref{eq:equationsLSTM}.

@@

\begin{equation}\label{eq:numcc_network-opt}
    (N \cdot K_G + 6) + 27  = 33 + N\cdot K_G
\end{equation}
which is only 13 clock cycles more than a fully-parallel, naïve architecture. For instance, an $N=32$ neuron network would
require 320 DSP slices, while this new architecture use 160, at the expense of 13 more clock cycles.

@@

Since the previous values are the time needed for a \emph{single} forward propagation, to know how many forward propagations we can perform \emph{per second},
we only need to invert the previous values. These values are presented in Figure~\ref{fig:Mclass-psec}. While the $N=8$ and $K_G=2$ network is able to
perform around 3.15 million predictions per second, the Python model can only output around 14 thousand predictions, which is a very significant result that
proves the relevance of this implementation.

@@

\smallskip
\emph{\textbf{This work is financed by the ERDF – European Regional Development Fund through the Operational Programme for Competitiveness and Internationalisation - COMPETE 2020 Programme within project \guillemotleft POCI-01-0145-FEDER-006961\guillemotright, and by National Funds through the FCT – Fundação para a Ciência e a Tecnologia (Portuguese Foundation for Science and Technology) as part of project \guillemotleft UID/EEA/50014/2013\guillemotright.}}

>>>>

Hardware platforms could achieve a considerable speed-up over software implementations, which would prove
useful for high data throughput systems, where the calculation overhead is critical and limits the performance.
Furthermore, a hardware implementation can even benefit \emph{offline} Deep Learning tasks by providing the results
of a given experiment faster than in a regular CPU, increasing scientific productivity.

@@

This article begins by explaining what are LSTM networks  in Section~\ref{sec:lstmnn}, followed by a quick overview of their state
of the art in Section~\ref{sec:relwork}. Section~\ref{sec:proparch} outlines the proposed
hardware architecture and its main constituent modules, and the performance and synthesis
results are reported in Section~\ref{sec:results}. Concluding remarks are given in
Section~\ref{sec:concl}.

@@

\begin{enumerate}
    \item Multiply matrix $\mb{W}_*$ by the input vector $\mb{x}^{(t)}$
    \item Multiply matrix $\mb{R}_*$ by the previous layer output vector $\mb{y}^{(t-1)}$
    \item Sum the bias vector $\mb{b}_*$ to the remaining matrix-vector dot product results.
\end{enumerate}
Assuming that the network size $N$ is always larger than the input size $M$, if we use the matrix-vector dot product units of Section~\ref{sec:proprarch_dot}, the multiplication
in task 1 takes approximately $K_G\cdot M$ cycles and the one in task 2 takes $K_G\cdot N$ cycles. This way, tasks 1 and 2 can be performed in parallel, and we can use the extra
time that task 2 takes, relative to task 1,  to perform task 3, and sum the bias vector to the output of task 1, whose result is ready by that time. The module is triggered by a
\verb+beginCalc+ input signal that activates the internal state machine, and outputs a \verb+dataReady+ signal that informs the network that the calculations have been concluded. 
Taking into account the internal state-machine and that the internal datapath is pipelined, the exact number
of clock cycles this module takes to produce an output is $6+K_G\cdot N$.

@@

Furthermore, we can avoid a naive translation of the Equations~\ref{eq:equationsLSTM}, which would replicate unnecessary resources (such as elementwise multipliers
and activation function calculators) and require more area to save a negligible number of clock cycles, by noting that one of the operands is the output
of a $\tanh(\mb{x})$ block and the other of a $\sigma(\mb{x})$, and they are then multiplied, elementwise, together. Instead of replicating
these `$\tanh$-$\sigma$-$(\cdot)$wise' structures, we use a \emph{single one} and choose the operand according to the state that the network is currently in. The
issue about the elementwise multiplier for $\mb{c}^{(t)}$, which does not use the $\tanh$ activation function, can be solved by adding another multiplexer that chooses
between the output of the $\tanh(\mb{x})$ module or the signal $\mb{c}^{(t-1)}$. These ideas resulted in the LSTM network design of Figure~\ref{fig:network-opt}, which
is mathematically equivalent to Equations~\ref{eq:equationsLSTM}.

@@

\begin{equation}\label{eq:numcc_network-opt}
    (N \cdot K_G + 6) + 27  = 33 + N\cdot K_G
\end{equation}
which is only 13 clock cycles more than a fully-parallel, naïve architecture. For instance, an $N=32$ neuron network would
require 320 DSP slices, while this new architecture uses 160, at the expense of 13 more clock cycles.

@@

To know how many forward propagations we can perform \emph{per second},
we only need to invert the previous values. These values are presented in Figure~\ref{fig:Mclass-psec}. While the $N=8$ and $K_G=2$ network is able to
perform around 3.15 million predictions per second, the Python model can only output around 14 thousand predictions, which is a very significant result that
proves the relevance of this implementation.

@@

\section*{Acknowledgements}
This work is financed by the ERDF – European Regional Development Fund through the Operational Programme for Competitiveness and Internationalisation - COMPETE 2020 Programme within project \guillemotleft POCI-01-0145-FEDER-006961\guillemotright, and by National Funds through the FCT – Fundação para a Ciência e a Tecnologia (Portuguese Foundation for Science and Technology) as part of project \guillemotleft UID/EEA/50014/2013\guillemotright.
